\begin{abstract}
Android applications commonly combine an Android Runtime (ART) managed heap with
native C/C++ allocations.  ART exposes managed-heap state, whereas Linux accounts
resident pages and memory pressure; neither view alone identifies how both heaps
contribute to an application's memory trajectory.  We present Heap Memory
Safeguard (HMS), a cross-layer controller design that combines resident occupancy
and recent allocation velocity and uses a two-threshold policy to request bounded,
memory-cgroup-scoped reclaim.  This paper evaluates the control policy in a
deterministic simulation artifact; it does not claim an Android, ART, or kernel
deployment.  Four synthetic workload families exercise bursty media, gaming,
inference, and multitasking patterns.  The artifact supports seeded replay,
parameter sweeps, and baseline comparisons, and reports simulated heap variability,
reclaim latency, garbage-collection pause, and peak resident footprint.  These
results establish feasibility and expose controller trade-offs, but they are not
device measurements.  We therefore treat Android integration, comparison with
LMKD, \texttt{memory.reclaim}, DAMON/DAMOS, and MGLRU, and validation with raw
Perfetto traces as required future work rather than completed contributions.
\end{abstract}

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10011007.10011006.10011050.10011051</concept_id>
       <concept_desc>Software and its engineering~Allocation / deallocation strategies</concept_desc>
       <concept_significance>500</concept_significance>
   </concept>
   <concept>
       <concept_id>10011007.10011074.10011099</concept_id>
       <concept_desc>Software and its engineering~Operating systems</concept_desc>
       <concept_significance>300</concept_significance>
   </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Allocation / deallocation strategies}
\ccsdesc[300]{Software and its engineering~Operating systems}

\keywords{memory management, cross-layer control, Android, memory cgroups,
simulation, reproducible artifact}
